---
title: "Beta Diversity"
output: html_notebook
---

This Notebook

Load libraries
```{r}
source("../General/rmb_functions.R")
source("../General/parameters.R")
library(vegan)
library(broom)
library(contrast)
library(cowplot)
library(tidyverse)
```

Load data
```{r}
wuf <- readRDS("../Data/wuf.RDS")
map <- readRDS("../Data/drought_map.RDS")
```

Remove samples used to train RF model
Generate a column that treats time point as a discrete variable rather than continuous (relevant for testing the effect of treatment at each time point)
```{r}
map <- filter(map, Treatment2 != "WC_TRN")
map$TimeFctr <- as.factor(map$Time)
```


Maybe
```{r}
time.age <- map %>% group_by(Time, Age) %>% count() %>% select(-n)
```

PERMANOVA
```{r}
wuf <- wuf[match(map$SampleID, rownames(wuf)), match(map$SampleID, colnames(wuf))]
pmanova <- adonis(as.dist(wuf) ~ Compartment * Time * Treatment,  data = map, permutations = 9999)
pmanova
```

PCoA
```{r}
pcoa.axes <- pcoa_axes(wuf, map)
pcoa.eig <- pcoa_eigval(wuf, map)

all.pcoa.p <- pcoa.axes %>% 
  mutate(Compartment = fct_recode(Compartment,
                                  "Rhizosphere" = "RS",
                                  "Endosphere" = "ES")) %>% 
  mutate(Compartment = fct_relevel(Compartment, "Rhizosphere")) %>% 
  ggplot(aes(Axis.1, Axis.2)) +
  geom_point(aes(color = Compartment), size = 2, shape = 16) +
  xlab(paste("PCo1 (", pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", pcoa.eig$Eigval[2], "%)", sep = "")) +
  scale_color_manual(values = cmp.pal) +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = c(0.25,0.8))

all.pcoa.p
```

Within Compartment Distance
```{r}
dist.tidy <- wuf %>% 
  as.matrix() %>% 
  as.data.frame() %>% 
  mutate(SampleID.x = row.names(.)) %>% 
  gather(key = "SampleID.y", value = "Distance", -SampleID.x) %>% 
  filter(!is.na(Distance)) %>% 
  filter(Distance > 0) %>% 
  inner_join(select(map, SampleID, Compartment, Treatment, Time, CollectionDate, Age, WaterStatus), by = c("SampleID.x" = "SampleID")) %>% 
  inner_join(select(map, SampleID, Compartment, Treatment, Time, CollectionDate, Age, WaterStatus), by = c("SampleID.y" = "SampleID"))


dist.within <- dist.tidy %>% 
  filter(Compartment.x == Compartment.y) %>% 
  filter(Treatment.x == Treatment.y) %>% 
  filter(Time.x == Time.y)
  
cmp.box <- dist.within %>% 
  mutate(Compartment = fct_relevel(Compartment.x, "RS")) %>% 
  mutate(Compartment = fct_recode(Compartment, Rhizosphere = "RS", Endosphere = "ES")) %>% 
  ggplot(aes(Compartment, Distance, color = Compartment)) +
  geom_boxplot(size = 1) +
  geom_rug() +
  ylab("Within-Compartment\nDistance") +
  xlab("Compartment") +
  scale_color_manual(values = cmp.pal) +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = "none")

cmp.box
```

```{r}
plot_grid(all.pcoa.p,
          cmp.box,
          nrow = 1,
          align = "v",
          labels = c("A", "B"),
          label_size = 20)
```

Subset rhizosphere and endosphere datasets
```{r}
wuf <- as.matrix(wuf)

rs.map <- filter(map, Compartment == "RS")
rs.wuf <- wuf[match(rs.map$SampleID, row.names(wuf)), match(rs.map$SampleID, colnames(wuf))]

es.map <- filter(map, Compartment == "ES")
es.wuf <- wuf[match(es.map$SampleID, row.names(wuf)), match(es.map$SampleID, colnames(wuf))]
```

PERMANOVA per compartment
```{r}
rs.pmanova <- adonis(as.dist(rs.wuf) ~ Time * Treatment,  data = rs.map, permutations = 9999)
rs.pmanova

es.pmanova <- adonis(as.dist(es.wuf) ~ Time * Treatment,  data = es.map, permutations = 9999)
es.pmanova
```

PCoA per compartment
```{r}
rs.pcoa.axes <- pcoa_axes(rs.wuf, rs.map)
rs.pcoa.eig <- pcoa_eigval(rs.wuf, rs.map)

es.pcoa.axes <- pcoa_axes(es.wuf, es.map)
es.pcoa.eig <- pcoa_eigval(es.wuf, es.map)
```

Plot PCoA
```{r}
rs.pc.time <- rs.pcoa.axes %>% 
  ggplot(aes(Axis.1, Axis.2, color = Age)) +
  geom_point(size = 2) +
  scale_color_viridis_c(name = "Plant Age\n(days)") +
  xlab(paste("PCo1 (", rs.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", rs.pcoa.eig$Eigval[2], "%)", sep = "")) +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = "top",
        legend.title.align = 1)

rs.pc.trt <- rs.pcoa.axes %>% 
  mutate(Treatment = fct_recode(Treatment,
                                "DS1" = "D1",
                                "DS2" = "D2",
                                "DS3" = "D3")) %>% 
  mutate(Treatment = fct_relevel(Treatment, "WC")) %>% 
  ggplot(aes(Axis.1, Axis.2, color = Treatment)) +
  geom_point(size = 2) +
  scale_color_manual(values = trt.pal,
                     guide = guide_legend(title.position = "top",
                                          title.hjust = 0.5)) +
  xlab(paste("PCo1 (", rs.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", rs.pcoa.eig$Eigval[2], "%)", sep = "")) +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = "top")

es.pc.time <- es.pcoa.axes %>% 
  ggplot(aes(Axis.1, Axis.2, color = Age)) +
  geom_point(size = 2) +
  scale_color_viridis_c() +
  xlab(paste("PCo1 (", es.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", es.pcoa.eig$Eigval[2], "%)", sep = "")) +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = "none")

es.pc.trt <- es.pcoa.axes %>% 
  ggplot(aes(Axis.1, Axis.2, color = Treatment)) +
  geom_point(size = 2) +
  scale_color_manual(values = trt.pal) +
  xlab(paste("PCo1 (", es.pcoa.eig$Eigval[1], "%)", sep = "")) +
  ylab(paste("PCo2 (", es.pcoa.eig$Eigval[2], "%)", sep = "")) +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = "none")
```

```{r}
plot_grid(all.pcoa.p, cmp.box,
          get_legend(rs.pc.time), get_legend(rs.pc.trt),
          rs.pc.time + theme(legend.position = "none"), rs.pc.trt + theme(legend.position = "none"),
          es.pc.time, es.pc.trt,
          nrow = 4,
          align = "vh",
          axis = "l",
          rel_heights = c(3,1,3,3),
          labels = c("A","B", NA, NA, "C", "D", "E", "F"),
          label_size = 20)
```


Functions to test the effect of drought treatment on principal coordinates
```{r}
run_lm <- function(df) {
  lm(value ~ Treatment * TimeFctr, data = df)
}

run_anova <- function(fit) {
  anova(fit) %>% tidy()
}

get_contrasts <- function(fit) {
  contrasts <- vector(mode = "list")

  for(i in 1:13) {
    i <- as.character(i)
    print(i)
    contrasts[[paste("D1", i, sep = ".")]] <-  contrast(fit, list(Treatment = "WC", TimeFctr = i), list(Treatment = "D1", TimeFctr = i))[1:8] %>% as.data.frame()
    contrasts[[paste("D2", i, sep = ".")]] <-  contrast(fit, list(Treatment = "WC", TimeFctr = i), list(Treatment = "D2", TimeFctr = i))[1:8] %>% as.data.frame()
    contrasts[[paste("D3", i, sep = ".")]] <-  contrast(fit, list(Treatment = "WC", TimeFctr = i), list(Treatment = "D3", TimeFctr = i))[1:8] %>% as.data.frame()
  }
  
  contrasts <- plyr::ldply(contrasts, function(x) x)
  contrasts <- contrasts %>% 
    separate(.id, c("Treatment", "Time")) %>% 
    mutate(Time = as.integer(Time),
           Significance = case_when(
             Pvalue >= 0.05 ~ "NS",
             Pvalue < 0.05 & Pvalue >= 0.01 ~ "<0.05",
             Pvalue < 0.01 & Pvalue >= 0.001 ~ "<0.01",
             Pvalue < 0.001 ~ "<0.001"))
}
```

Run lms 
```{r}
rs.pcoa.lm <- rs.pcoa.axes %>% 
  select(Axis.1:Axis.5, SampleID:TimeFctr) %>% 
  gather(key = "Axis", value = "value", Axis.1:Axis.5) %>% 
  group_by(Axis) %>% 
  nest() %>% 
  mutate(lm = map(data, run_lm),
         anova = map(lm, run_anova),
         contrasts = map(lm, get_contrasts))

es.pcoa.lm <- es.pcoa.axes %>% 
  select(Axis.1:Axis.5, SampleID:TimeFctr) %>% 
  gather(key = "Axis", value = "value", Axis.1:Axis.5) %>% 
  group_by(Axis) %>% 
  nest() %>% 
  mutate(lm = map(data, run_lm),
         anova = map(lm, run_anova),
         contrasts = map(lm, get_contrasts))
```

Make a data frame to transform time points to ages
```{r}
time.age <- map %>% group_by(Time, Age) %>% count() %>% select(-n)
```


```{r}
rs.means <- rs.pcoa.axes %>% 
  group_by(Treatment, Age) %>% 
  mutate(Mean = mean(Axis.1)) %>% 
  ungroup()

rs.pc <- rs.pcoa.axes %>% 
  ggplot(aes(Age, Axis.1)) +
  geom_vline(xintercept = ws.line, linetype = 3, color = "gray20", size = 0.5) +
  geom_point(aes(color = Treatment), alpha = 1, shape = 1, size = 2) +
  geom_point(aes(0,0), alpha = 0) +
  geom_line(data = rs.means, aes(y = Mean, color = Treatment), size = 1) +
  xlab("Plant Age (days)") +
  ylab(paste("Rhizopshere\nPCo1 (", rs.pcoa.eig$Eigval[1], "%)", sep = "")) +
  scale_fill_manual(values = trt.pal) +
  scale_color_manual(values = trt.pal) +
  scale_x_continuous(breaks = seq(0,140,20)) +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 15))

rs.sig <- rs.pcoa.lm %>% 
  unnest(contrasts) %>% 
  filter(Axis == "Axis.1") %>% 
  ungroup() %>% 
  mutate(p.adj = p.adjust(Pvalue, method = "fdr")) %>% 
  inner_join(time.age, by = "Time") %>% 
  ggplot(aes(Age, Treatment)) +
  geom_point(aes(color = Treatment, alpha = p.adj < 0.05), shape = "*", size = 7) +
  geom_point(aes(0,1), alpha = 0) +
  scale_y_discrete(limits = c("D3", "D2", "D1")) +
  scale_x_continuous(breaks = seq(0,140,20)) +
  scale_color_manual(values = trt.pal[-1]) +
  scale_alpha_manual(values = c(0,1)) +
  theme_void() +
  theme(legend.position = "none")

rs.plot <- plot_grid(rs.sig, rs.pc, ncol = 1, rel_heights = c(3,17), align = "v")

rs.plot
```

```{r}
es.means <- es.pcoa.axes %>% 
  group_by(Treatment, Age) %>% 
  mutate(Mean = mean(Axis.1)) %>% 
  ungroup()

es.pc <- es.pcoa.axes %>% 
  ggplot(aes(Age, Axis.1)) +
  geom_vline(xintercept = ws.line, linetype = 3, color = "gray20", size = 0.5) +
  geom_point(aes(color = Treatment), shape = 1, size = 2, alpha = 1) +
  geom_point(aes(0,0), alpha = 0) +
  geom_line(data = es.means, aes(y = Mean, color = Treatment), size = 1) +
  xlab("Plant Age (days)") +
  ylab(paste("Endosphere\nPCo1 (", es.pcoa.eig$Eigval[1], "%)", sep = "")) +
  scale_fill_manual(values = trt.pal) +
  scale_color_manual(values = trt.pal) +
  scale_x_continuous(breaks = seq(0,140,20)) +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 15))


es.sig <- es.pcoa.lm %>% 
  unnest(contrasts) %>% 
  filter(Axis == "Axis.1") %>% 
  ungroup() %>% 
  mutate(p.adj = p.adjust(Pvalue, method = "fdr")) %>% 
  inner_join(time.age, by = "Time") %>% 
  ggplot(aes(Age, Treatment)) +
  geom_point(aes(color = Treatment, alpha = p.adj < 0.05), shape = "*", size = 7) +
  geom_point(aes(0,1), alpha = 0) +
  scale_y_discrete(limits = c("D3", "D2", "D1")) +
  scale_x_continuous(breaks = seq(0,140,20)) +
  scale_color_manual(values = trt.pal[-1]) +
  scale_alpha_manual(values = c(0,1)) +
  theme_void() +
  theme(legend.position = "none")

es.plot <- plot_grid(es.sig, es.pc, ncol = 1, rel_heights = c(3,17), align = "v")

es.plot
```

```{r}
soil <- map %>% 
  filter(Treatment2 != "WC_TRN") %>% 
  filter(Compartment == "RS") %>% 
  mutate(TimeFctr = as.factor(Time))
```

Run an anova to check for significant effects
```{r}
soil.lm <- lm(PercentMoisture ~ Treatment * TimeFctr, data = soil)
soil.aov <- anova(soil.lm) %>% 
  tidy()
soil.aov
```

```{r}
contrasts <- vector(mode = "list")

for(i in 1:13) {
  i <- as.character(i)
  print(i)
  contrasts[[paste("D1", i, sep = ".")]] <-  contrast(soil.lm, list(Treatment = "WC", TimeFctr = i), list(Treatment = "D1", TimeFctr = i))[1:8] %>% as.data.frame()
  contrasts[[paste("D2", i, sep = ".")]] <-  contrast(soil.lm, list(Treatment = "WC", TimeFctr = i), list(Treatment = "D2", TimeFctr = i))[1:8] %>% as.data.frame()
  contrasts[[paste("D3", i, sep = ".")]] <-  contrast(soil.lm, list(Treatment = "WC", TimeFctr = i), list(Treatment = "D3", TimeFctr = i))[1:8] %>% as.data.frame()
}

contrasts <- plyr::ldply(contrasts, function(x) x)
contrasts <- contrasts %>% 
  separate(.id, c("Treatment", "Time")) %>% 
  mutate(Time = as.integer(Time),
         Significance = case_when(
           Pvalue >= 0.05 ~ "NS",
           Pvalue < 0.05 & Pvalue >= 0.01 ~ "<0.05",
           Pvalue < 0.01 & Pvalue >= 0.001 ~ "<0.01",
           Pvalue < 0.001 ~ "<0.001")) 
```

```{r}
soil.means <- map %>% 
  group_by(Age, Treatment, WaterStatus) %>% 
  summarise(MeanPM = mean(PercentMoisture))

soil.p <- soil %>% 
  ggplot(aes(Age, PercentMoisture)) +
  geom_vline(xintercept = ws.line, linetype = 3, color = "gray20", size = 0.5) +
  geom_point(aes(color = Treatment), shape = 1, size = 2, alpha = 1) +
  geom_point(aes(0,20), alpha = 0) +
  geom_line(data = soil.means, aes(Age, MeanPM, color = Treatment), size= 1) +
  scale_color_manual(values = trt.pal) +
  scale_x_continuous(breaks = seq(0,140,20)) +
  scale_shape_manual(values = ws.shp) +
  ylab("Soil Moisture\n(% weight)") +
  xlab("Plant Age (days)") +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 15))

soil.p

sig.p <- contrasts %>% 
  ungroup() %>% 
  mutate(p.adj = p.adjust(Pvalue, method = "fdr")) %>% 
  inner_join(time.age, by = "Time") %>% 
  ggplot(aes(Age, Treatment)) +
  geom_point(aes(color = Treatment, alpha = Pvalue < 0.05), shape = "*", size = 7) +
  geom_point(aes(0,1), alpha = 0) +
  scale_y_discrete(limits = c("D3", "D2", "D1")) +
  scale_x_continuous(breaks = seq(0,140,20)) +
  scale_color_manual(values = trt.pal[-1]) +
  scale_alpha_manual(values = c(0,1)) +
  xlab("Plant age (days)") +
  #xlim(0,NA) +
  theme_void() +
  theme(legend.position = "none") 

sig.p

soil.plot <- plot_grid(sig.p, soil.p, ncol = 1, rel_heights = c(3,17), align = "v")
soil.plot
```

```{r}
xpt.ds <-  map %>% 
  mutate(Treatment = case_when(Treatment == "WC" ~ "WC",
                                    Treatment == "D1" ~ "DS1",
                                    Treatment == "D2" ~ "DS2",
                                    Treatment == "D3" ~ "DS3")) %>% 
  mutate(Treatment = fct_relevel(Treatment, "WC")) 

xpt.p <- xpt.ds %>% 
  filter(WaterStatus != "S") %>% 
  ggplot(aes(PrevAge, Treatment)) +
  geom_segment(aes(xend = Age, yend = Treatment, color = Treatment), size = 3, linetype = 1) +
  geom_segment(data = filter(xpt.ds, WaterStatus == "S"), aes( xend = Age, yend = Treatment, color = Treatment), linetype = 2) +
  geom_point(aes(Age, 4.5), shape = 6, size = 2) +
  geom_vline(xintercept = ws.line, linetype = 3, color = "gray20", size = 0.5) +
  xlab("Plant Age (days)") +
  scale_color_manual(values = trt.pal) +
  scale_y_discrete(limits = c("DS3", "DS2", "DS1", "WC")) +
  scale_x_continuous(breaks = seq(0,140,20),
                     position = "bottom") +
  scale_linetype_manual(values = c(1,3)) +
  theme_classic() +
  theme(text = element_text(size = 15),
        axis.line.y = element_blank(),
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        legend.position = "none")


xpt.p
```

```{r}
#386:1073

plot_grid(xpt.p,
          sig.p, soil.p, #+ theme(axis.title.x = element_blank()),
          rs.sig, rs.pc, #  + theme(axis.title.x = element_blank()),
          es.sig,es.pc,
          ncol = 1,
          align = "v",
          rel_heights = c(10,
                          3,17,
                          3,17,
                          3,17),
          labels = c("A", "B", NA, "C", NA, "D", NA),
          label_size = 20)
```

```{r}
whole.map <- readRDS("../Data/drought_map.RDS")

tubs.p <- whole.map %>% 
  mutate(Treatment = str_replace(Treatment, "D", "DS")) %>% 
  mutate(Treatment = fct_relevel(Treatment, "WC")) %>% 
  mutate(Treatment2 = str_replace(Treatment2, "D", "DS")) %>% 
  ggplot(aes(Column, Row, fill = Treatment)) +
  geom_tile(color = "white", size = 2) +
  geom_text(aes(label = Treatment2), color = "white", size = 5) +
  scale_fill_manual(values = trt.pal) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.title = element_blank(),
        text = element_text(size = 20),
        panel.border = element_blank(),
        panel.grid = element_blank(),
        axis.ticks = element_blank())

tubs.p
```

